### 2010-03-13 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: merge Conor's comments in. (first pass)

Our proposal is based on induction-recursion~\cite{dybjer:general-ir,
  dybjer:axiom-ir, dybjer:ir-initial-algebra, dybjer:iir}. However, in
the context of simple data-types, the full power of
induction-recursion is not necessary. Therefore, we will present a
stripped-down version. As we extend the scope of our universe of
data-types, our presentation will diverge from induction-recursion. We
shall compare both approaches in Section~\ref{sec:discussion}

@@

In dependently-typed language, $\Sigma$-types can be read in two
rather different ways. This duality is actually reflected in the
notation we can find in the literature, depending on the sensibility
of the author. Hence, the type $\SigmaTy{A}{B}$ is sometimes written
$\Sigma_{x : A} (B x)$. This notation stresses the fact that
$\Sigma$-types are a generalisation of sums over arbitrary
arities. When simply-typed languages have finite sums,
dependently-typed languages have sums of indexed by any set.

On the other hand, $\SigmaTy{A}{B}$ is sometimes written
$\SIGMA{\V{x}}{A} (B x)$. Under this perspective, $\Sigma$-types can
be read as a generalisation of products, where the second component
can depend on the first one. When simply-typed languages pack data
into non-dependent tuples, dependently-types languages tolerate this
non-dependent usage. However, they also give the rather novel ability
for data to influence further data.

@@

In the simply-typed world, the essence of data-type definition can be
summed up by the term ``sum-of-product''. A data-type is defined by a
finite sum, choice, of constructors, each of them composed by a
product, a tuple, of arguments. Therefore, the grammar of data-types
definition in simply-typed languages, such as Haskell or OCaml, follow
this form.

@@

For our sigma-of-sigmas to be known, we crucially need the ability to
name them. To this end, we will use a standard technique in
dependently-typed programming: we construct an ad-hoc universe. This
technique dates back to Martin-L\"of definition of type
theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
programming technique~\cite{who?} \note{I would need some other
  example of universe construction for practical purposes}. We refer
the reader to Agda's tutorial~\cite{norell:agda-tutorial} for a
pedagogical presentation of universe construction.

@@

Hence, we propose to embed inductive types as a universe in our
dependent type theory, the universe of \emph{descriptions}
(Figure~\ref{fig:desc_universe}). As expected, the code of this
universe mimics the standard grammar of data-types definitions in
simply-typed languages. Hence, we have a $\DSigma{}{}$ code,
interpreted as a $\Sigma$-type, to build the
sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
contains no useful payload. The functoriality of the data-types is
introduced by $\DIndx$. When we tie the knot with a fix-point, the
hole left open by $\DIndx$ will be turned into a standard recursive
argument. We notice that this functoriality appears in the type of
$\descop{D}{}$ itself, for a given data-type definition $D$. This
corresponds to the morphism part of the functor described by $D$.

@@

To give some intuition on this universe of descriptions, we now turn
to some examples. For obvious pedagogical reasons, we will manually
build these descriptions. However, it should be clear that, in
practice, these definitions can be automatically constructed an
Haskell-like $\data$ definition.

@@

Our first example is the natural numbers, or rather its carrier
functor. Our code is presented in the high-level expression language
of Section~\ref{sec:type-propagation}. The translation back to the raw
terms is laborious but should not pose any difficulty. The code is the
following:

@@

\[    \NatD\: Z = 1 + Z    \]

@@

With a small change to the definition of $\NatD$, we obtain the
carrier of lists:

@@

The modification consists in turning the $\NatSuc{}$ into a proper
$\ListCons$ taking an argument in X followed by an inductive
argument. In this case, we use $\DSigma{}{}$ in its product
interpretation: we pack an element of $X$ together with the recursive
argument. Easily, one sees that this code actually defines the
following functor:

\[    \ListD\: X Z = 1 + X \times Z     \]

@@

Finally, we are not limited to one recursive argument. This is
demonstrated by our description of a binary tree functor below:

@@

Again, we are at one evolutionary step away from $\ListD$. However,
instead of single appeal to the induction code, we call it twice. The
interpretation of this code corresponds to the following functor:

\[    \TreeD\: X Z = 1 + X \times Z \times Z     \]

@@

From the examples above, we observe that data-types are defined by a
$\DSigma$ which first argument is a finite set of constructor. The
descriptions fitting into this pattern are called \emph{tagged}
description. Formally, we have:

@@

\[
 \TagDesc \mapsto \SIGMA{\V{E}}{\EnumU} \spi{\V{E}}{\LAM{\_} \Desc}
\]

Again, this is a clear reminiscence of the ``sum-of-products'' style:
we have a finite sum of constructors. Of course, every description can
be expressed into this style, by using a singleton as single
constructor.

@@

So far, for convenience, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its code, $\DSigma{}{}$, $\DIndx{}$,
and $\DUnit$ as type formers, extending the basic type theory. Hence,
in an implementation of this type theory, we would have extended it
with these constructors and their typing rules. Although it makes no
conceptual difference, the code of description should rather be read
as a specification. We promise the existence of such objects,
satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
implementing the specification.

@@

The first step consists in turning $T$ into a predicate
$\Bhab{P}{\Mu{D} \To \Set}$, hence introducing a dependency on the
descriptions. Consequently, we strengthen the algebra $\descop{D}{T}
\To T$ by a similar process. We start with turning the arrow into a
dependently arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
further analysis.

Then, we have to translate the notion of algebra in our dependent
setting. This is achieved by the $\All{D}{X}{P}{x}$ predicate that,
intuitively, states that $\Bhab{P}{X \To \Set}$ holds everywhere in
the sub-structures of $x$. Based on that knowledge, we ought to be
able to prove that $P$ holds for $\Con{xs}$. We close our elimination
principle by taking an argument $x$ in $\Mu{D}$ and claiming that the
motive $P$ is realised on $x$.

@@

At this stage, we are still left with implementing $\F{All}$ and the
$\F{induction}$ principle. For space reason, we refer the reader to
our technical report for the actual code. Having this dependent
catamorphism, we are empowered with the ability to $\emph{compute}$
over descriptions. Note that this induction principle is the first
manifestation of a generic operation over descriptions. Thanks to
$\F{induction}$, any data-type we define automatically comes equipped
with an induction principle.

@@

As for finite sets previously, these symptoms clearly indicate that we
suffer from too much type information. Just as for finite, the cure
lies in rationalising this information. Let us first tackle the issue
of data-type declaration: unsurprisingly, the grammar of data-types we
are used can be straightforwardly translated to the
``sigmas-of-sigmas'' paradigm. Hence, a data-type definition is turned
into the corresponding description. Then, at the level of definitions,
we associate the name of the data-type to the fix-point of the
description. The induction principle as well as other generic
operations follow for free.

Doing so, we seal data-type definition behind a convenient
abstraction. We take advantage of this abstraction barrier to tamper
the second issue. Namely, our user would rather be presented a name
rather that a fix-point of a signature functor. Rather surprisingly,
it is sufficient to \emph{label} the $\Mu{}$ constructor with the
user-provided name. When pretty-printing this object, we just expose
the label, instead of the definition in full.

>>>>

Our proposal is inspired by, but less involved than, Dybjer and
Setzer's closed formulation of
induction-recursion~\cite{dybjer:general-ir, dybjer:axiom-ir,
  dybjer:ir-initial-algebra, dybjer:iir}. In the context of simple
data-types, the full power of induction-recursion is not
necessary. Therefore, we adopt a stripped-down version of it. As we
extend the scope of our universe of data-types, our presentation will
diverge from induction-recursion. We shall compare both approaches in
Section~\ref{sec:discussion}

@@

In dependently-typed language, $\Sigma$-types can be read in two
rather different ways. This duality is actually reflected in the
notation we can find in the literature, depending on the sensibility
of the author. Hence, the type $\SigmaTy{A}{B}$ is sometimes written
$\Sigma_{x : A} (B\: x)$. This notation stresses the fact that
$\Sigma$-types are a generalisation of sums over arbitrary
arities. When simply-typed languages have finite sums,
dependently-typed languages have sums indexed by any set.

On the other hand, $\SigmaTy{A}{B}$ is sometimes written
$\SIGMA{\V{x}}{A} (B\: x)$. Under this perspective, $\Sigma$-types can
be read as a generalisation of products, where the second component
can depend on the first one. When simply-typed languages pack data
into non-dependent tuples, dependently-types languages tolerate this
non-dependent usage. However, they also give the rather novel ability
for data to influence further data.

@@

In the simply-typed world, the essence of data-type definition can be
summed up by the term ``sum-of-product''. A data-type is defined by a
finite sum -- a choice -- of constructors, each of them is composed by
a product -- a tuple -- of arguments. Therefore, the grammar of
data-types definition in simply-typed languages, such as Haskell or
OCaml, follow this form.

@@

For our sigma-of-sigmas to be known, we crucially need the ability to
name them. To this end, we will use a standard technique in
dependently-typed programming: we construct an ad-hoc universe. This
technique dates back to Martin-L\"of definition of type
theory~\cite{martin-lof:itt}. Since then, it has been fruitful as a
programming technique~\cite{benke:universe-generic-prog, whoelse?}
\note{I would need some other example of universe construction for
  practical purposes}. We refer the reader to Agda's
tutorial~\cite{norell:agda-tutorial} for a pedagogical presentation of
universe construction.

@@

Hence, we propose to embed inductive types as a universe in our
dependent type theory, the universe of \emph{descriptions}
(Figure~\ref{fig:desc_universe}). As expected, the code of this
universe mimics the standard grammar of data-types definitions in
simply-typed languages. Hence, we have a $\DSigma{}{}$ code,
interpreted as a $\Sigma$-type, to build the
sigmas-of-sigmas. Descriptions are terminated by $\DUnit$, which
contains no useful payload. The functoriality of the data-types is
introduced by $\DIndx$. When we tie the knot with a fix-point, the
hole left open by $\DIndx$ will be turned into a standard recursive
argument. We notice that this functoriality appears in the type of
$\descop{D}{}$ itself, for a given data-type definition $D$. This
corresponds to the object part of the functor described by $D$.

@@

To give some intuition on this universe of descriptions, we now turn
to some examples. For obvious pedagogical reasons, we will manually
build these descriptions. However, it should be clear that, in
practice, these definitions can be automatically constructed from an
Haskell-like $\data$ definition.

@@

Our first example is the natural numbers, or rather its signature
functor. Our code is presented in the high-level expression language
of Section~\ref{sec:type-propagation}. The translation back to the raw
terms is laborious but should not pose any difficulty. The code is the
following:

@@

\[    \NatD\: Z \mapsto 1 + Z    \]

@@

With a small change to the definition of $\NatD$, we obtain the
signature of lists:

@@

The modification consists in turning the $\NatSuc{}$ constructor into
a proper $\ListCons$ taking an argument in X followed by a recursive
argument. In this case, we use $\DSigma{}{}$ in its product
interpretation: we pack an element of $X$ together with the recursive
argument. Easily, one sees that this code actually defines the
following functor:

\[    \ListD\: X\: Z \mapsto 1 + X \times Z     \]

@@

Finally, we are not limited to one recursive argument. This is
demonstrated by our description of binary trees:

@@

Again, we are at one evolutionary step away from $\ListD$. However,
instead of a single call to the induction code, we call it twice. The
interpretation of this code corresponds to the following functor:

\[    \TreeD\: X\: Z \mapsto 1 + X \times Z \times Z     \]

@@

From the examples above, we observe that data-types are defined by a
$\DSigma$ which first argument is a finite set of constructors. The
descriptions fitting into this pattern are called \emph{tagged}
description. Formally, we have:

@@

\[
 \TagDesc \mapsto \SIGMA{\V{E}}{\EnumU} (\spi{\V{E}}{(\LAM{\_} \Desc)})
\]

Again, this is a clear reminiscence of the ``sum-of-products'' style:
we have a finite sum of constructors. Of course, every description can
be expressed into this style, using a singleton as single constructor
if necessary.

@@

So far, for convenience, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its code, $\DSigma{}{}$, $\DIndx{}$,
and $\DUnit$ as type formers, extending the basic type theory. Hence,
in an implementation of this type theory, we need to extended it with
these constructors and their typing rules. Although it makes no
conceptual difference, the code of description should rather be read
as a specification. We promise the existence of such objects,
satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfil this promise, by actually
implementing the specification.

@@

The first step consists in turning $T$ into a predicate
$\Bhab{P}{\Mu{D} \To \Set}$, hence introducing a dependency on the
descriptions. Consequently, we strengthen the algebra $\descop{D}{T}
\To T$ by a similar process. We start with turning the arrow into a
dependent arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
further analysis.

Then, we have to translate the notion of algebra in our dependent
setting. This is achieved by the predicate $\All{D}{X}{P}{x}$ that,
intuitively, states that $\Bhab{P}{X \To \Set}$ holds everywhere in
$x$. Based on that knowledge, we ought to be able to prove that $P$
holds for $\Con{xs}$. We close our elimination principle by taking an
argument $x$ in $\Mu{D}$ and claiming that the motive $P$ is realised
on $x$.

@@

At this stage, we are still left with implementing this
$\F{induction}$ principle. For space reason, we refer the reader to
our technical report for the actual code, hence the proof of (weak)
initiality. Having this dependent catamorphism, we are empowered with
the ability to $\emph{compute}$ over descriptions. Note that this
induction principle is the first manifestation of a generic operation
over descriptions. Thanks to $\F{induction}$, any data-type we define
automatically comes equipped with an induction principle.

@@

As for finite sets previously, these symptoms clearly indicate that we
suffer from too much type information. Just as for finite sets, the
cure lies in rationalising this information. Let us first tackle the
issue of data-type declaration. The grammar of data-types we are used
to can be straightforwardly translated to the ``sigmas-of-sigmas''
paradigm. Hence, a data-type definition can be turned into the
corresponding description. Then, at the level of definitions, we
associate the name of the data-type to the fix-point of the
description. The induction principle as well as other generic
operations follow for free.

Doing so, we seal data-type definition behind a convenient
abstraction. We take advantage of this abstraction barrier to tamper
the second issue. Namely, our user would rather be presented a name
rather than a fix-point of a signature functor. Surprisingly, it is
sufficient to \emph{label} the $\Mu{}$ constructor with the
user-provided name. When pretty-printing this object, we just expose
the label, instead of the definition in full.

### 2010-03-13 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: merge Conor's comments in. (first pass)

Another interesting consequence concerns the $\spi{}{}$ operator: it
can actually be implemented with the generic $\F{induction}$
principle. From an implementation perspective, this means that we
remove an operator, hard-coded in the host language. This operator is
simply implemented in the dependently-typed target language, as any
other function.

@@

If not on the content, this modification had an effect on the
names. The type naming scheme of the type theory has condensed: named
type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are now
replaced by codes and their fix-point in the universe of
descriptions. Our next step is similar, in essence: we are going to
condense the entire naming scheme of data-types \emph{in itself}.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, $\DescD$ first consists in choosing
a constructor. Constructors correspond to the code's cases: $\DUnit$,
$\DSigma$, and $\DIndx$. The reader will notice that the tagged
notation we have used so far for $\Desc$ constructors now fully make
sense: these were actually the tags we are defining.

@@

Note that $\DIndx$ is isomorphic to $\DHindx{\Unit}$. However, we
tolerate this duplication and do not replace $\DIndx$ by the more
general $\DHindx$. Indeed, unlike its counterpart, $DIndx$ is
first-order. Therefore, while both codes are \emph{extensionally} the
same, in most practical implementation, they would be dealt with
rather differently. While we can definitional equality to cope with
first order objects, the functional presentation introduced by
$\DHindx$ is unlikely to be amenable to a purely definitional
treatment.

Equipped with $\DHindx$, we can describe the extended code of our
universe of data-types, as shown in Figure~\ref{fig:desc-levitate}.
The $\DUnit$ and $\DIndx$ cases remains unchanged, as expected. We
achieve the description of the $\DSigma$ case, by a simple appeal to
the higher-order induction on $S$. The $\DHindx$ case consists in
packing the $H$ in $\Set$ with a recursive argument.

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.\ref{fig:desc-levitate}). Or rather, it is made invisible by a
careful usage of type propagation. Indeed, let us elaborate this term
down to the low-level type theory. The $\EnumT [ \ldots ]$ construct
elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $f$ is pushed into the type
$\EnumT{e} \To \Desc$. This corresponds to a finite function set we
have implemented in Figure~\ref{fig:type-checking}.

@@

\[
\PLAM{x}{(\EnumT{e})} \switch{e}{\PLAM{\_} \Desc}{\pi^f}{x}
\]

@@

If we cannot make it happen, we simply do not make it happen. We
perfectly know what the motive ought to be. So, straightforwardly, we
extend the type theory with two special-purpose operators:

\[
\begin{array}{@{}ll}
%% piD
\green{\pi_D} : & \PITEL{\V{e}}{\EnumU} \To \Set \\
%% switchD
\F{switchD} : & \PITEL{\V{e}}{\EnumU}               
                \PITEL{\V{b}}{\spiD{e}}
                \PITEL{\V{x}}{\EnumT{e}} \To \Desc
\end{array}
\]

The entire work of the magician stands here, in these two
extensions. One could be worried by the availability of such
operations in our type theory. Indeed, it would be inconvenient for
the user to have to decide between these two variants. In fact, we
give no choice to the user, hence no inconvenience. Just as the
$\F{switch}$ operator was sealed off by the type propagation
machinery, $\F{switchD}$ is not provided to the user. We simply extend
the type-checking rules (Fig.~\ref{fig:type-checking}) with the
following inference rule:

@@

This concludes our levitation work. Beyond the beauty of the exercise,
there is also several practical outcomes. First of all, this exercise
reveals that the $\Desc$ universe is just plain data. Just as any
piece of data, it can therefore be inspected and manipulated. However,
unlike any data, it is expressed in the $\Desc$ universe. As a
consequence, it is equipped, for free, with an induction
principle. So, our ability to inspect and program with $\Desc$ is not
restricted to the meta-language: we now have all the necessary
equipment in the target language to \emph{program} over data-types. In
this setting, generic programming is just about \emph{programming}.

@@

Before moving on some examples of generic programming, we have to
carry out some maintenance work. The extension of $\Desc$ with
$\DHindx$ have introduced some duplication of functionality. In order
to remove this clutter, we will factor out these commonalities. This
results in a new definition of the $\Desc$ universe. This
reorganisation is strictly equivalent to the previous one
(Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
self-descriptive. We refer the reader to our companion technical
report for details.

@@

\begin{wstructure}
<- Deriving generic functions
    <- Taking a Desc and computing a function
        <- Desc comes equipped with an induction principle
        -> Ability to compute more functions from it
            -> More generic functions
    <- Inspecting data-types
        <- All described by a Desc code
        -> Ability to explore the code
            <- Desc equipped with an induction principle
            -> Build new objects based on that structure
\end{wstructure}

With this example, we have shown how we can derive a generic
operation, the catamorphism, from a pre-existing generic operation,
the induction principle. This has been made possible by our ability to
manipulate descriptions as first-class objects: the catamorphism is,
basically, a function mapping a $\Desc$ to a data-type specific
operation. Having the ability to program on $\Desc$ is the key to
define such function. The induction principle of $\Desc$, inherited by
levitation, plays a central role here: without an induction principle,
it would be impossible to program on $\Desc$.

@@

We simply add a constructor, $\DVar{}$, and define its argument to be
a $\DSigma{X}{\Unit}$, that is an element of $X$. We keep $E$ and $D$
as they were, hence leaving the functor untouched. The interpretation
of this definition convince us that we have built the functor
signature of the free monad. The fix-point operation ties the knot and
gives us the full-blown free monad construction.

@@

Filling the $???$ sub-goal with $\apply\: D\: X\: Y\: \sigma$ finishes
the implementation. To sum up, we have implemented the free monad
construction for an arbitrary tagged description. This gives our user
the ability, for any data-type, to extend it with a notion of
variable. Then, with have equipped this structure with the
corresponding monadic operation, ``bind'' and ``return''. Whereas the
``return'' is a trivial variable former, the ``bind'' operation
corresponds to the substitution of variables by terms. We have shown
how to implement it, for any free monad.

@@

With the free monad construction, we have seen two kind of generic
operations. Firstly, we have derived a new data-structure from another
one: we make the free monad from its underlying functor. To do so, we
crucially rely on the fact that data-types are nothing but codes. We
are therefore entitled to modify it and, in this case, extend
it. Extending a data-type might give raise to a more
structured object, as was the case here.

So, secondly, we have equipped this new data-type with its inherent
structure: the bind and return operations. We have been able to build
them as generic functions. Here, we rely on the ability to compute
over descriptions, thanks to the induction principle.

>>>>

Another interesting consequence concerns the $\spi{}{}$ operator: it
can actually be implemented with the generic $\F{induction}$
principle. From an implementation perspective, this means that we
remove an operator hard-coded in the host language. This operator is
simply implemented in the dependently-typed target language, as any
other function.

@@

If not on the content, this modification had an effect on the
names. The type naming scheme of the type theory has condensed: named
type formers ($\EnumU$) and constructors ($\NilE$ and $\ConsE$) are
now replaced by codes and their fix-point in the universe of
descriptions. In essence, our next step is similar: we are going to
condense the entire naming scheme of data-types \emph{in itself}.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the data-types we have seen so far, we first have to choose a
constructor. Constructors correspond to the code's cases: $\DUnit$,
$\DSigma$, and $\DIndx$. The reader will notice that the tagged
notation we have used so far for $\Desc$ constructors now fully make
sense: these were actually the tags we are defining. \note{This
  sentence badly needs a rephrasing.}

@@

Note that $\DIndx$ is isomorphic to $\DHindx{\Unit}$. However, we
tolerate this duplication and do not replace $\DIndx$ by the more
general $\DHindx$. Indeed, unlike its counterpart, $\DIndx$ is
first-order. Therefore, while both codes are \emph{extensionally} the
same, in most practical implementation they would be dealt with rather
differently. While definitional equality can cope with first order
objects, the functional presentation introduced by $\DHindx$ is
unlikely to be amenable to a purely definitional treatment.

Equipped with $\DHindx$, we can describe the extended code of our
universe of data-types, as shown in Figure~\ref{fig:desc-levitate}.
The $\DUnit$ and $\DIndx$ cases remains unchanged, as expected. We
successfully describe the $\DSigma$ case, by a simple appeal to the
higher-order induction on $S$. The $\DHindx$ case consists in packing
the $H$ in $\Set$ with a recursive argument.

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us elaborate this term
down to the low-level type theory. The $\EnumT [ \ldots ]$ construct
elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $f$ is pushed into the type
$\EnumT{e} \To \Desc$. This corresponds to a finite function
definition we have implemented in Figure~\ref{fig:type-checking}.

@@

\[
\PLAM{x}{(\EnumT{e})} \switch{e}{\LAM{\_} \Mu{\DescD}}{\pi^f}{x}
\]

@@

If we cannot make it happen, we simply do not make it happen. We
perfectly know what the motive ought to be. Consequently, we extend
the type theory with a special-purpose operator:

\[
\begin{array}{@{}ll}
%% switchD
\F{switchD} : & \PITEL{\V{e}}{\EnumU}               
                \PITEL{\V{b}}{\spi{e}{\LAM{\_} \Desc}}
                \PITEL{\V{x}}{\EnumT{e}} \To \Desc
\end{array}
\]

The entire work of the magician stands here, in this extension. One
could be worried by the availability of such operations in our type
theory. Indeed, it would be inconvenient for the user to have to
decide between two variants. In fact, we give no choice to the user,
hence no inconvenience. Just as the $\F{switch}$ operator was sealed
off by the type propagation machinery, $\F{switchD}$ is not provided
to the user. We simply extend the type-checking rules
(Fig.~\ref{fig:type-checking}) with the following inference rule:

@@

This concludes our levitation work. Beyond the beauty of the exercise,
there are several practical outcomes. First of all, this exercise
reveals that the $\Desc$ universe is just plain data. Just as any
piece of data, it can therefore be inspected and manipulated. However,
unlike any data, it is expressed in the $\Desc$ universe. As a
consequence, it is equipped, for free, with an induction
principle. So, our ability to inspect and program with $\Desc$ is not
restricted to the meta-language: we now have all the necessary
equipment in the target language to \emph{program} over data-types. In
this setting, generic programming is just about \emph{programming}.

@@

Before moving on some examples of generic programming, we have to
carry out some maintenance work. The extension of $\Desc$ with
$\DHindx$ have introduced some duplication of functionality. In order
to remove this clutter, we will factor out these commonalities. This
results in a new definition of the $\Desc$ universe. This
reorganisation is strictly equivalent to the previous one
(Fig.~\ref{fig:hindx_desc}). Just as the previous version, it is also
self-descriptive. We refer the reader to the companion technical
report for details.

@@

\begin{wstructure}
<- Deriving generic functions
    <- Taking a Desc and computing a function
        <- Desc comes equipped with an induction principle
        -> Ability to compute more functions from it
            -> More generic functions
    <- Inspecting data-types
        <- All described byu a Desc code
        -> Ability to explore the code
            <- Desc equipped with an induction principle
            -> Build new objects based on that structure
\end{wstructure}

With this example, we have shown how we can derive a generic
operation, the catamorphism, from a pre-existing generic operation,
the induction principle. This has been made possible by our ability to
manipulate descriptions as first-class objects: the catamorphism is,
basically, a function mapping a $\Desc$ to a data-type specific
operation. Having the ability to program on $\Desc$ is the key to
define such function. The induction principle of $\Desc$ -- inherited
by levitation -- plays a central role here: without an induction
principle, it would be impossible to program on $\Desc$.

@@

We simply add a constructor, $\DVar{}$, and define its argument to be
a $\DSigma{X}{\Unit}$, that is an element of $X$. We keep $E$ and $D$
as they were, hence leaving the functor unchanged. The interpretation
of this definition convince us that we have built the functor
signature of the free monad. The fix-point operation ties the knot and
gives us the full-blown free monad construction.

@@

Filling the $???$ sub-goal with $\apply\: D\: X\: Y\: \sigma$ finishes
the implementation. To sum up, we have implemented the free monad
construction for an arbitrary tagged description. This gives our user
the ability, for any data-type, to extend it with a notion of
variable. Then, we have equipped this structure with the
corresponding monadic operation, ``bind'' and ``return''. Whereas the
``return'' is a trivial variable former, the ``bind'' operation
corresponds to the substitution of variables by terms. We have shown
how to implement it, for any free monad.

@@

With the free monad construction, we have seen two kind of generic
operations. Firstly, we have derived a new data-structure from another
one: we make the free monad from its underlying functor. To do so, we
crucially rely on the fact that data-types are nothing but codes. We
are therefore entitled to modify this code and, in this case, extend
it. Extending a data-type might give rise to a more structured
object, as was the case here.

So, secondly, we have equipped this new data-type with its inherent
structure: the bind and return operations. We have been able to build
them as generic functions. Here, we rely on our ability to compute
over descriptions, thanks to the induction principle.
